Well-quasi-orderings on word languages

### 2025-01-13 src/amalgam.tex
adding proofs for uniformly recurrent words, end

\begin{definition}
    An \intro{amalgamation system}
    is a tuple $(\Sigma, R, \canrun, E)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    $\canrun \colon R \to (\Sigma \uplus \set{\cansep})^*$ is a 
    function computing a \emph{canonical decomposition} of a run,
    and $E$ describes the so-called \emph{admissible embeddings} between runs: If $\rho$ and $\sigma$ are runs from $R$, then $E(\rho, \sigma)$ is a subset of the subword embeddings between $\canrun[\rho]$ and $\canrun[\sigma]$. We write $\rho \intro*\runleq \sigma$ if $E(\rho, \sigma)$ is non-empty. If we want to refer to a specific embedding $f \in E(\rho, \sigma)$, we also write $\rho \runleq_f \sigma$.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\intro*\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{$(R, E)$ Forms a Category.}
            For all $\rho, \sigma, \tau \in R$,
            $\mathrm{id} \in E(\rho,\rho)$,
            and whenever $f \in E(\rho,\sigma)$ and $g \in E(\sigma,\tau)$,
            then $g \circ f \in E(\rho,\tau)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \runleq)$ is a well-quasi-ordered set.
        \item \emph{Concatenative Amalgamation.}
            Let $\rho_0, \rho_1, \rho_2$ be runs 
            with $\rho_0 \runleq_f \rho_1$ 
            and $\rho_0 \runleq_g \rho_2$.
            Then for all $0 \leq i \leq \card{\canrun(\rho_0)}$,
            there exists a run $\rho_3 \in R$ 
            and embeddings $\rho_1 \runleq_{g'} \rho_3$
            and $\rho_2 \runleq_{f'} \rho_3$ 
            satisfying two conditions:
            (a) $g' \circ f = f' \circ g$ (we write $h$ for this composition) and
            (b) for every $0 \leq j \leq \card{\canrun[\rho_0]}$, 
            the gap word $\GapWord{h}{j}$
            is either $\GapWord{f}{j} \GapWord{g}{j}$
            or $\GapWord{h}{j} = \GapWord{g}{j} \GapWord{f}{j}$. 
            Specifically, for $i$ we may fix $\GapWord{h}{i} = \GapWord{f}{i} \GapWord{g}{i}$.
            We refer to \cref{amalgamation-runs:fig} for an illustration 
            of this property.
    \end{enumerate}

@@

    Indeed, assume that
    $L$ is \kl{downwards closed} and included in a finite union of sets of the form
    $\InfPeriodChain{x} u \InfPeriodChain{y}$ where $x,y,u$ are possibly empty words.
    We can assume without loss of generality that
    for every $n$, $x^n u y^n$ is in $L$, otherwise, we have a bound on the maximal $n$ such that
    $x^n u y^n$ is in $L$, and we can increase the number of languages in the union, replacing $x$ or $y$
    with the empty word as necessary.
    Let us write $L' \defined \bigcup_{i = 1}^k x_i^* u_i y_i^*$. Then, $L'
    \subseteq L$ by construction. Furthermore, $L \subseteq \dwset{L'}$, also
    by construction. Finally, we conclude that $L = \dwset{L'}$ because $L$ is
    \kl{downwards closed}. Now, because $L'$ is a \kl{regular language}, and 
    \kl{regular languges} are closed under \kl{downwards closure}, we conclude
    that $L$ is a \kl{regular language}.
\end{proof}

>>>>

\begin{definition}
    An \intro{amalgamation system}
    is a tuple $(\Sigma, R, \canrun, E)$ where
    $\Sigma$ is a finite alphabet,
    $R$ is a set of so-called \emph{runs},
    $\canrun \colon R \to (\Sigma \uplus \set{\cansep})^*$ is a 
    function computing a \intro{canonical decomposition} of a run,
    and $E$ describes the so-called \intro{admissible embeddings} between runs: If $\rho$ and $\sigma$ are runs from $R$, then $E(\rho, \sigma)$ is a subset of the subword embeddings between $\canrun[\rho]$ and $\canrun[\sigma]$. We write $\rho \intro*\runleq \sigma$ if $E(\rho, \sigma)$ is non-empty. If we want to refer to a specific embedding $f \in E(\rho, \sigma)$, we also write $\rho \runleq_f \sigma$.
    Given a run $r \in R$, and $i \in \range[0]{\card{\canrun(r)}}$, 
    the \intro{gap language} of $r$ at position $i$ is $\intro*\GapLanguage{r}{i} \defined
    \setof{\GapWord{f}{i}}{\exists s \in R. \exists f \in E(r,s) }$.
    An \kl{amalgamation system} furthermore satisfies the following 
    properties:
    \begin{enumerate}
        \item \emph{$(R, E)$ Forms a Category.}
            For all $\rho, \sigma, \tau \in R$,
            $\mathrm{id} \in E(\rho,\rho)$,
            and whenever $f \in E(\rho,\sigma)$ and $g \in E(\sigma,\tau)$,
            then $g \circ f \in E(\rho,\tau)$.
        \item \emph{Well-Quasi-Ordered System.}
            $(R, \runleq)$ is a well-quasi-ordered set.
        \item \emph{Concatenative Amalgamation.}
            Let $\rho_0, \rho_1, \rho_2$ be runs 
            with $\rho_0 \runleq_f \rho_1$ 
            and $\rho_0 \runleq_g \rho_2$.
            Then for all $0 \leq i \leq \card{\canrun(\rho_0)}$,
            there exists a run $\rho_3 \in R$ 
            and embeddings $\rho_1 \runleq_{g'} \rho_3$
            and $\rho_2 \runleq_{f'} \rho_3$ 
            satisfying two conditions:
            (a) $g' \circ f = f' \circ g$ (we write $h$ for this composition) and
            (b) for every $0 \leq j \leq \card{\canrun[\rho_0]}$, 
            the gap word $\GapWord{h}{j}$
            is either $\GapWord{f}{j} \GapWord{g}{j}$
            or $\GapWord{h}{j} = \GapWord{g}{j} \GapWord{f}{j}$. 
            Specifically, for $i$ we may fix $\GapWord{h}{i} = \GapWord{f}{i} \GapWord{g}{i}$.
            We refer to \cref{amalgamation-runs:fig} for an illustration 
            of this property.
    \end{enumerate}

@@

    Indeed, assume that
    $L$ is \kl{downwards closed} and included in a finite union of sets of the form
    $\InfPeriodChain{x} u \InfPeriodChain{y}$ where $x,y,u$ are possibly empty words.
    We can assume without loss of generality that
    for every $n$, $x^n u y^n$ is in $L$, otherwise, we have a bound on the maximal $n$ such that
    $x^n u y^n$ is in $L$, and we can increase the number of languages in the union, replacing $x$ or $y$
    with the empty word as necessary.
    Let us write $L' \defined \bigcup_{i = 1}^k x_i^* u_i y_i^*$. Then, $L'
    \subseteq L$ by construction. Furthermore, $L \subseteq \dwset{L'}$, also
    by construction. Finally, we conclude that $L = \dwset{L'}$ because $L$ is
    \kl{downwards closed}. Now, because $L'$ is a \kl{regular language}, and 
    \kl{regular languages} are closed under \kl{downwards closure}, we conclude
    that $L$ is a \kl{regular language}.
\end{proof}

### 2025-01-13 src/infixes.tex
adding proofs for uniformly recurrent words, end

@@

This motivated the following theorem, which states that for \kl{downwards
closed} languages, the \kl{ordinal invariants} are relatively small.

@@

\AP A subset $I \subseteq X$ is \intro{directed} if, for every $x,y \in I$,
there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl{directed} for the \kl{order relation} $\leq$. In
our case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix
relation} is the set of finite \kl{infixes} of a finite or bi-infinte word $w
\in \Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word.

@@

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}$. Now, let us build the sequence $\seqof{u_i}$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl{directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.

@@

\begin{lemma}
    \label{small-ordinal-invariants:lem}
    Let $w \in \Sigma^\Nat$ be an \kl{ultimately uniformly recurrent} word.
    Then, the set of finite infixes of $w$ has \kl{ordinal width}
    less than $2 \cdot \omegaOrd$.
\end{lemma}
\begin{proof}
    \todo[inline]{prove this}
\end{proof}

\begin{lemma}
    \label{from-bi-to-single:lem}
    Let $w \in \Sigma^\Rel$ be a bi-infinite word. Then, the set of finite
    infixes of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation} if
    and only if $w_+$ and $w_-$ are two \kl{ultimately uniformly recurrent}
    words. In this case, the \kl{ordinal width}
    of the set of finite infixes of $w$ is less than $3 \cdot \omegaOrd$.
\end{lemma}

Before proving \cref{small-ordinal-invariants:thm}, let us restrict 
our attention to a simpler case:

The \cref{small-ordinal-invariants:thm} is a direct consequence of the
following remarks.

\begin{itemize}
    \item If $L$ is \kl{wqo} and \kl{downwards closed}, then $L = \bigcup_{i = 1}^n L_i$ where $L_i$
        is an \kl{order ideal}.
    \item If $L$ is an \kl{order ideal} that is \kl{well-quasi-ordered}, then
        $L$ is the set of infixes of some \emph{bi-infinite} word $w \in \Sigma^{\Rel}$.
    \item The set of infixes of a \emph{bi-infinte} word can be understood 
        as the set of \kl{infixes} of two infinite words $u,v \in \Sigma^{\Nat}$.
    \item The set of infixes of an infinite word $w \in \Sigma^{\Nat}$ is 
        \kl{well-quasi-ordered} if and only if $w$ is \kl{ultimately uniformly recurrent},
        that is, if there exists an infinite suffix of $w$ that is \kl{uniformly recurrent}.
    \item The ordinal invariants of a \kl{uniformly recurrent} word are small.
\end{itemize}

Note that being \kl{uniformly recurrent} still leaves open a lot of potential behaviours,
for instance, Sturmian words are \kl{uniformly recurrent}, but are not \kl{automatic}.
\todo{Find citations for this.}

\begin{theorem}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}, and if so, compute the associated \kl{ordinal
    invariants}.
\end{theorem}
\begin{proof}
    One can check if $w$ is \kl{ultimately uniformly recurrent} by checking if
    $\exists N_0 \in \Nat, \dots$. Because $w$ is \kl{automatic}, this is decidable.
    Now, to decide whether $w$ has width $\omegaOrd$, then one can check if $w$ is 
    \kl{ultimately periodic}. If not, then $w$ has width $\omegaOrd$.
    If it is \kl{ultimately periodic}, one can compute the smallest size of
    the period, which gives us the width.
    Using the width, and because we know the height is at most $\omegaOrd$, we can
    compute the maximal order type.
\end{proof}

>>>>

\begin{corollary}
    \label{ordinal-invariants-bounded:cor}
    Let $L$ be a \kl{bounded language} of $\Sigma^*$
    that is \kl{well-quasi-ordered} by the \kl{infix relation}.
    Then, the \kl{ordinal width} of $L$ is finite,
    its \kl{ordinal height} is at most $\omega$,
    and its \kl{maximal order type} is at most $\omega$.
\end{corollary}

@@

We prove if the upcoming theorem that the status of the \kl{Thue-Morse
sequence} is actually representative of most \kl{downwards closed} languages
for the \kl{infix relation}. Indeed, we show that the \kl{ordinal invariants}
of such languages are relatively small, and the proof of this fact relies on
the connection between \kl{well-quasi-ordered} languages and \kl{uniformly
recurrent} words.

@@

\AP A subset $I \subseteq X$ is \intro(subset){directed} if, for every $x,y \in
I$, there exists $z \in I$ such that $x \leq z$ and $y \leq z$. Given a
\kl{well-quasi-order} $(X, \leq)$, one can always decompose $X$ into a finite
union of \intro{order ideals}, that is, non-empty sets $I \subseteq X$ that are
\kl{downwards closed} and \kl(subset){directed} for the relation $\leq$. In our
case, a \kl{well-quasi-ordered} \kl{order ideal} for the \kl{infix relation} is
the set of finite \kl{infixes} of a finite or bi-infinte word $w \in
\Sigma^\Nat \cup \Sigma^\Rel $ (\cref{bi-infinite:lem}). Therefore, we will
first study languages that are described as the set of finite \kl{infixes} of
some (bi-)infinite word (\cref{ultimately-uniformly-recurrent:lem}). To that
end, let us introduce the notation $\intro*\infset{w}$ for the set of finite
\kl{infixes} of a (possibly infinite or bi-infinite) word $w$.

@@

    Because the alphabet $\Sigma$ is finite, we can enumerate the words of $L$
    as $\seqof{w_i}$. Now, let us build the sequence $\seqof{u_i}$ by induction
    as follows: $u_0 = w_0$, and $u_{i+1}$ is a word that contains $u_i$ and
    $w_i$, which exists in $L$ because $L$ is \kl(subset){directed}. Since $L$ is
    \kl{well-quasi-ordered}, one can extract an infinite set of indices $I
    \subseteq \Nat$ such that $u_i \infleq u_{j}$ for all $i \leq j \in I$.

@@

\begin{lemma}
    \label{small-ordinal-invariants:lem}
    Let $w \in \Sigma^\Nat$ be an \kl{ultimately uniformly recurrent} word.
    Then, the set of finite infixes of $w$ has \kl{ordinal width}
    less than $2 \cdot \omegaOrd$.
\end{lemma}
\begin{proof}
    Let $N_0$ be a bound such that $w_{\geq N_0}$ is \kl{uniformly recurrent}.
    Let us write $\infset{w}$ the set of finite infixes of $w$.
    We prove that $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.
    Let $u_1 \infleq w$ be a finite word.

    If $u_1$ is an \kl{infix} of $w_{\geq N_0}$, then there exists $k \geq 1$
    such that $u_1$ is an \kl{infix} of every word of size at least $k$. In
    particular, there is finite bound on the length of every sequence of
    incomparable elements starting with $u_1$. We conclude in particular that
    $\infset{w} \setminus \upset{u_1}$ has a finite \kl{ordinal width}.

    Otherwise, $u_1$ can only be found \emph{before} $N_0$. In this case, we
    consider a second element of a \kl{bad sequence} $u_2 \infleq w$, which is
    incomparable with $u_1$ for the \kl{infix relation}. If $u_2$ is an
    \kl{infix} of $w_{\geq N_0}$, then we can conclude as before. Otherwise,
    notice that $u_1$ and $u_2$ cannot start at the same position in $w$
    (because they are incomparable). Continuing this argument, we conclude that
    there are at most $N_0$ elements starting before $N_0$
    at the start of any sequence of
    incomparable elements in $\infset{w}$. We conclude that
    $\oWidth{\infset{w}} \leq \omegaOrd + N_0$.
\end{proof}

Let us briefly discuss the tightness of the bound in
\cref{small-ordinal-invariants:lem}. The \kl{Thue-Morse sequence} over a binary
alphabet $\set{a,b}$ has \kl{ordinal width} $\omegaOrd$. Given a number $N_0
\in \Nat$, one can construct an arbitrarily long \kl{antichain} of words for
the \kl{infix relation} by using a new letter $c$. When concatenating this
(finite) antichain as a prefix of the \kl{Thue-Morse sequence}, one obtains a
new (infinite) word $w$. It is clear that the \kl{ordinal width} of
$\infset{w}$ is now at least $\omegaOrd + N_0$. Hence, the bound $\oWidth{X} <
2 \cdot \omegaOrd$ is tight.

The behaviour of bi-infinite words is very similar to the infinite ones. Given
a bi-infinite word $w \in \Sigma^{\Rel}$, we can consider $w_+ \in \Sigma^\Nat$
and $w_- \in \Sigma^\Nat$ the two infinite words obtained as follows: for all
$i \in \Nat$, $(w_+)_i = w(i)$ and $(w_-)_i = w(-i)$. Note that the two share
the letter at position $0$.

\begin{lemma}
    \label{from-bi-to-single:lem}
    Let $w \in \Sigma^\Rel$ be a bi-infinite word. Then, the set of finite
    infixes of $w$ is \kl{well-quasi-ordered} for the \kl{infix relation} if
    and only if $w_+$ and $w_-$ are two \kl{ultimately uniformly recurrent}
    words. In this case, the \kl{ordinal width}
    of the set of finite infixes of $w$ is less than $3 \cdot \omegaOrd$.
\end{lemma}
\begin{proof}
    Assume that $w_+$ and $w_-$ are \kl{ultimately uniformly recurrent}. Let us
    write $\infset{w}$ the set of finite infixes of $w$. Consider an infinite
    sequence of words $\seqof{u_i}[i \in \Nat]$ in $\infset{w}$. If there is an
    infinite subsequence of words that are all in $\infset{w_+}$, then there
    exists an increasing pair of indices $i < j$ such that $u_i \infleq
    u_j$ because \cref{uniformly-recurrent:lem} applies to $w_+$. Similarly, if
    there is an infinite subsequence of words that are all in $\infset{w_-}$,
    then there exists an increasing pair of indices $i < j$ such that $u_i
    \infleq u_j$ because \cref{uniformly-recurrent:lem} applies to $w_-$ (and
    the \kl{infix relation} is compatible with mirroring). Otherwise, one can
    assume without loss of generality that all words in the sequence have a
    starting position in $w_-$ and an ending position in $w_+$. In this case,
    let us write $(k_i,l_i) \in \Nat^2$ the pair of indices such that $u_i$ is the infix
    of $w$ that starts at position $-k_i$ of $w$ (i.e., $k_i$ of $w_-$) and
    ends at position $l_i$ of $w$ (i.e., $l_i$ of $w_+$).
    Because $\Nat^2$ is a \kl{well-quasi-ordering} with the product ordering,
    there exists $i < j$ such that $k_i \leq k_j$ and $l_i \leq l_j$, 
    in particular, $u_i \infleq u_j$. We have proven that every infinite
    sequence of words in $\infset{w}$ is \kl(wqo){good}, hence $\infset{w}$ is
    \kl{well-quasi-ordered}.

    Conversely, assume that $\infset{w}$ is \kl{well-quasi-ordered}. In
    particular, $\infset{w_+} \subseteq \infset{w}$ is \kl{well-quasi-ordered}.
    Similarly, $\infset{w_-}$ is \kl{well-quasi-ordered} because the \kl{infix
    relation} is compatible with mirroring. Applying
    \cref{ultimately-uniformly-recurrent:lem}, we conclude that both are
    \kl{ultimately uniformly recurrent} words.

    To obtain the upper bound of $3 \cdot \omegaOrd$, we can consider the same
    argument as for \cref{small-ordinal-invariants:lem}. We let $N_0$ be such
    that $w_{\geq N_0}$ and $(w_-)_{\geq N_0}$ are \kl{uniformly recurrent}
    words. In any sequence of incomparable elements of $\infset{w}$, there are
    less than $N_0^2$ elements that are found in $(w_{< N_0})_{> -N_0}$. Then,
    one has to pick a finite \kl{infix} in either $w_{\geq N_0}$ or $w_{\leq
    -N_0}$. Because of \cref{small-ordinal-invariants:lem}, any sequence of
    incomparable elements of these two infinite words has length bounded based
    on the choice of the first element of that sequence. This means that the
    \kl{ordinal width} of $\infset{w}$ is at most $\omegaOrd + \omegaOrd +
    N_0^2$. We conclude that $\oWidth{\infset{w}} < 3 \cdot \omegaOrd$.
\end{proof}

As for \cref{small-ordinal-invariants:lem}, let us briefly discuss
that the bound in \cref{from-bi-to-single:lem} is tight. Indeed, one can
construct a bi-infinite word $w$ by concatenating a reversed \kl{Thue-Morse
sequence} on a binary alphabet $\set{a,b}$, a finite antichain of arbitrarily
large size over a distinct alphabet $\set{c,d}$, and then a \kl{Thue-Morse
sequence} on a binary alphabet $\set{e,f}$. The \kl{ordinal width} of the set
of \kl{infixes} of $w$ is then at least $2 \cdot \omegaOrd + K$, where $K$ is the
size of the chosen antichain.

We are now ready to conclude the proof of \cref{small-ordinal-invariants:thm}.
\begin{proofof}{small-ordinal-invariants:thm}

    It is always true that the \kl{ordinal height} of a language over a finite
    alphabet is at most $\omegaOrd$. Let us now consider a
    \kl{well-quasi-ordered} language $L$ that is \kl{downwards closed} for the
    \kl{infix relation}. Because it is a \kl{well-quasi-ordered} set, it can be
    written as a finite union of \kl{order ideals} $L = \bigcup_{i = 1}^n L_i$.

    For every such \kl(order){ideal} $L_i$, we can apply
    \cref{bi-infinite:lem}, and conclude that $L_i$ is the set of finite
    \kl{infixes} of a finite, infinite or bi-infinite word $w_i$. We can then
    directly conclude that $\oWidth{L_i}$ less than $\omegaOrd$ (in the case of
    a finite word), less than $2 \cdot \omegaOrd$ (in the case of an infinite
    word thanks to \cref{small-ordinal-invariants:lem}), or less than $3 \cdot
    \omegaOrd$ (in the case of a bi-infinite word, thanks to
    \cref{from-bi-to-single:lem}). In any case,
    we have the bound $\oWidth{L_i} < 3 \cdot \omegaOrd$.

    Now, $\oWidth{L} \leq \sum_{i = 1}^n \oWidth{L_i} < 3 \cdot \omegaOrd <
    \omegaOrd^2$. We conclude that the \kl{ordinal width} of $L$ is less than
    $\omegaOrd^2$. Finally, the inequality $\oType{L} \leq \oWidth{L} \oComProd
    \oHeight{L} < \omega \oComProd \omega^2 = \omegaOrd^3$ allows us to conclude.
\end{proofof}

\AP As we have demonstrated, infinite (or bi-infinite words) can be used to
represent languages that are \kl{well-quasi-ordered} for the \kl{infix
relation} by considering their set of finite \kl{infixes}. We can leverage the
theory of \kl{automatic sequences}, that is tailored to study the regularity of
infinite words to build a decision procedure for \kl{well-quasi-ordered}
languages for the \kl{infix relation}. A sequence $w \in \Sigma^\Nat$ is
\intro(sequence){automatic} whenever there exists a base $b \in \Nat$ and a
finite automaton that can compute the $i$-th letter of $w$ given as input the
number $i$ in base $b$. The \kl{Thue-Morse sequence} is an example of an
\kl{automatic sequence}.

\begin{theorem}
    Given an \kl{automatic sequence} $w \in \Sigma^{\Nat}$, one can decide
    whether the set of \kl{infixes} of $w$ is \kl{well-quasi-ordered} for the
    \kl{infix relation}.
\end{theorem}
\begin{proof}
    Because of \cref{ultimately-uniformly-recurrent:lem}, it suffices
    to decide whether $w$ is \kl{ultimately uniformly recurrent}.
    We can rewrite this as a question on the \kl{automatic sequence} $w$
    as follows:
    \begin{align*}
        &\exists N_0,                   &   \text{ultimately} \\
        &\forall i_s \geq N_0,          &   \text{for every infix (start) } u \\
        &\forall i_e > i_s,             &   \text{for every infix (end) }   u \\
        &\exists k \geq 1,              &   \text{there exists a bound} \\
        &\forall j_s \geq N_0,          &   \text{for every other infix (start) } v \\
        &\forall j_e \geq j_s + k,      &   \text{of size at least $k$} \\
        &\exists l \geq 0,              &   \text{there exists a position in } v \\
        &\forall 0 \leq m < i_e - i_s,  &   \text{where } u \text{ can be found} \\
        &j_s + m + l < j_e \land
        w(i_s + m) = w(j_s + m + l) \quad .
    \end{align*}
    Because $w$ is computable by a finite automaton, one can reduce the above
    formula to a regular language, for which it suffices to check emptiness, which
    is decidable.
\end{proof}

Note that being \kl{uniformly recurrent} still leaves open a lot of potential
behaviours, for instance, Sturmian words are \kl{uniformly recurrent}, but are
not \kl(sequence){automatic}. Interestingly enough, the \kl{ordinal invariants}
of the \kl{automatic sequences} are not smaller than the ones of \kl{ultimately
uniformly recurrent} words: this contrasts with the case of languages that are
\kl{well-quasi-ordered} for the \kl{infix relation} in general, where the
hypothesis of being representable (by a \kl{regular language}, or an
\kl{amalgamation system}) will drastically reduce the \kl{ordinal invariants}
of said language (\cref{ordinal-invariants-bounded:cor}, \cref{infix-amalgamation:thm}).
